\section{Lecture 2: Basic Mathematics II}

\subsection{Summation Operator}

\begin{itemize}
    \item Consider a set $X = \{x_1, x_2, ..., x_n\}$
    \begin{itemize}
        \item $\sum\limits_{i = 1}^{n} x_i = x_1 + x_2 + ... + x_n$
        \begin{itemize}
            \item ``The sum of $x_i$, over the range from $i=l$ through $i = n$.''
        \end{itemize}
    \end{itemize}
    \item Let $Y = \{y_1, y_2, ..., y_n\}$ 
    \begin{itemize}
        \item $\sum\limits_{i=1}^n x_iy_i = x_1y_1 + x_2y_2 + ... + x_ny_n$ 
        \begin{itemize}
            \item Here, we sum the outcome of $x_iy_i$ $n$ times.
        \end{itemize}
    \end{itemize}
    \item But, what if there is more than one summation operator and subscript?
    \vspace{-1em}
    \begin{align*}
        \Sigma_{j=1}^m \Sigma_{i=1}^n x_iy_j 
        & = (y_1x_1 + y_1x_2 + ... + y_1x_n)   \\
        & + (y_2x_1 + y_2x_2 + ... + y_2x_n) \\
        & + ... + (y_mx_1 + y_mx_2 + ... + y_mx_n)
    \end{align*}
    We have $m$ parentheses, with $n$ $x$'s in each parenthesis. We end up summing each combination of $x_i$ and $y_j$.
    \item What if there is a constant, $c$? Drawing upon the \emph{distributive property}:
    \begin{itemize}
        \item $\sum\limits_{i = 1}^{n} cx_i = cx_1 + cx_2 + ... + cx_n = c\sum\limits_{i = 1}^{n} x_i$
    \end{itemize}
    \item The \emph{associative property} can also be applied to summation:
    \begin{itemize}
        \item $\sum\limits_{i=1}^n (x_i + y_i) = (x_1 + y_1) + (x_2 + y_2)
            + ... + (x_n + y_n) = \sum\limits_{i=1}^n x_i + \sum\limits_{i=1}^n y_i$
    \end{itemize}
\end{itemize}

\subsection{Product Operator}

\begin{itemize}
    \item $X = \{x_1, x_2, ..., x_n\}$ and $Y = \{y_1, y_2, ..., y_n\}$ 
    \item $\prod\limits_{i=1}^n x_i = x_1 \cdot x_2 \cdot ... \cdot x_n$
    \item $\prod\limits_{i=1}^n (x_i + y_i) = (x_1 + y_1)(x_2 + y_2) ...
        (x_n + y_n)$
    \begin{itemize}
        \item We can't split $(x_i + y_i)$ like in summation. Instead, we multiply $(x_i + y_i)$ repeatedly.
    \end{itemize}
    \item What about a constant? 
    \begin{itemize}
        \item $\prod\limits_{i = 1}^{n} cx_i = (cx_1)(cx_2) \cdot ... \cdot
            (cx_n) = c^n\prod\limits_{i = 1}^{n} x_i$
    \end{itemize}
    We can move $c$ to the front, but we have to exponentiate it to $c^n$.
\end{itemize}

\subsection{Factorials, Permutations, and Combinations}

Most of our quantitative coursework is about modeling probabilities, where: 

\begin{equation*}
    \text{probability} = \frac{\# \text{occurences}}{\# \text{possibilities}}
\end{equation*}

\noindent For both the numerator and denominator we are dealing with \emph{counting} the number of relevant outcomes. Factorials, permutations, and combinations are foundational concepts when it comes to counting. 

\begin{itemize}
    \item Some useful illustrations/properties:
    \begin{itemize}
        \item $0! = 1$ 
        \item $x! = x \cdot (x-1) \cdot (x-2) \cdot ... \cdot 0!$
        \item $2! = 2 \cdot 1$
        \item $3! = 3 \cdot 2 \cdot 1$
        \item $10! = 10 \cdot 9 \cdot 8 \cdot 7 \cdot 6 \cdot ... \cdot 1$ 
        \item If $n$ objects in boxes of size $x$ where \emph{order matters}, then the number of \emph{permutations} is:
        \begin{equation*}
            \frac{n!}{(n - x)}
        \end{equation*}
        \item If order does not matter, then the number of \emph{combinations} is:
        \begin{equation*}
            \frac{n!}{(n - x)! x!} = n C x = {n \choose x }
        \end{equation*}
        \begin{itemize}
            \item Pronounced: `$n$ choose $x$'
        \end{itemize}
    \end{itemize}
\end{itemize}

\subsection{Solving equations, inequalities, and for roots}

\begin{itemize}
    \item Solving an equation example:
    \begin{align*}
        3x + 4y + 8 & = 0 \\
        4y & = - (3x + 8) \\
        y & = \frac{- (3x + 8)}{4} \\
        y & = -\frac{3}{4}x - 2
    \end{align*}
    
    \noindent We can also write this answer as:
    
    \begin{equation*}
        \{(x,y) \in \mathbb{R} | y = - \frac{3}{4} x - 2 \}
    \end{equation*}
    
    \item Solving an inequality ($x > $, $x \geq y$, $y < x$, $y \leq x$) example:
    \begin{align*}
        -4y & > 2x + 12 \\
        y & < -\frac{2x}{4} - \frac{12}{4} \\
        y & < \frac{x}{2} - 3
    \end{align*}

    \noindent Note that dividing by a negative flips the sign

    \item Solving for a quadratic
    \begin{itemize}
        \item Quadratic formula:
        \begin{align*}
            & ax^2 + bx + c \\
            & x \in \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
        \end{align*}
        \noindent Example:
        \begin{align*}
            & 1.4x^2 + 3.7x + 1.1 = 0 \\
            & a = 1.4, b = 3.7, c = 1.1 \\
            & x = \frac{-3.7 \pm \sqrt{3.7^2 -
                4 \times 1.4 \times 1.1}}{2.8} \\
            & x = - 0.341 \text{ or } x = -2.301 
        \end{align*}
    \end{itemize}
\end{itemize}

\subsection{Logarithms}

\begin{itemize}
    \item If $y = a^x$, then we can rewrite as $\text{log}_a y = x$
    \item Examples: 
    \begin{itemize}
        \item $\text{log}_e e = 1$, because $e^1 = e$ 
        \item Solving for $x$: 
        \begin{itemize}
            \item If $8 = 2^x$, then $\text{log}_2 8 = x = 3$
        \end{itemize}
    \end{itemize}
    \item \textbf{Rules}
    \begin{itemize}
        \item $\text{log}(m + n) = \text{log}(m) + \text{log}(n)$ 
        \item $\text{log}(\frac{m}{n}) = \text{log}(m) - \text{log}(n)$
        \item $\text{log}(b^a) = a \text{log} b$
        \item $\text{log}b^a = (\text{log}_be)(\text{log}_ea)$ 
        \item $\text{log}b^a = \frac{1}{\text{log}_ab}$
    \end{itemize}
    \item \textbf{Visual}
    
    \resizebox{4in}{!}{
        \input{figures/logarithm.tex}
    }
    \item Note: $log_e(x) \equiv ln(x)$, pronounced the `natural logarithm' 
\end{itemize}